Time Series Analysis with Bandt \& Pompe Symbolization and a Test for White Noise in the Entropy-Complexity Plane

### 2021-12-19 Reports/JOURNAL - Confidence Regions/IT-ConfidenceRegionsR2.tex
Version R2

\fundinginfo{CNPq, Fapeal}

@@

Let ${\mathbbm A}_{D}$ (with $D \geq 2$ and $D \in {\mathbbm N}$) be the symmetric group of order $D!$ formed by all 
possible permutation of order $D$, and the symbol component vector 
${\bm \pi}^{(D)} = (\pi_1, \pi_2, \dots, \pi_D)$ so every element ${\bm \pi}^{(D)}$ is unique 
($\pi_j \neq \pi_k~\forall~j \neq k$). 
Consider for the time series ${\mathcal X} \equiv \{x_t\}_{t=1}^{T}$ its time delay embedding representation,
with embedding dimension $D \geq 2$ ($D \in {\mathbbm N}$) and time delay $\tau \geq 1$ ($\tau \in {\mathbbm N}$, also called ``embedding time''):
\begin{equation} 
{\mathbf X}^{(D,\tau)}_t =( x_t,x_{t+\tau},\dots,x_{t+(D-1)\tau} ) ,
\label{eq:time-delay}
\end{equation} 
for $t = 1,2,\dots,N$ with $N = T-(D-1) \tau$.
Then, the vector ${\mathbf X}^{(D,\tau)}_t$ can be mapped to a symbol ${\bm \pi}^{(D)} \in {\mathbbm A}_{D}$. 
This mapping should be defined in a way that preserves the desired relation between the elements 
$x_t  \in {\mathbf X}^{(D,\tau)}_t$, and all $t \in T$ that share this pattern (also called ``motif'') are mapped to the same 
${\bm \pi}^{(D)}$. 
The two most frequent ways to define the mapping ${\mathbf X}^{(D,\tau)} \mapsto {\bm \pi}^{(D)}$ are:  
\begin{enumerate}[label=\alph*)]
\item ordering the ranks of $x_t \in {\mathbf X}^{(D,\tau)}$ in chronological order 
(\textit{Rank Permutation}) or,
\item ordering the time indexes of $x_t \in {\mathbf X}^{(D,\tau)}$  
(\textit{Chronological Index Permutation}).
\end{enumerate}
See details in the work by \citet{BPRepeatedValuesChaos}.
Without loss of generality, in the following, we will use the latter.

@@

After computing all the symbols, one obtains the histogram of proportions $\bm h = (h(j))_{1\leq j\leq D!}$.
Such histogram estimates the (unknown, in general) probability distribution function of these patterns.
The next step into the characterization of the time series is computing descriptors from this histogram.

@@

The Jensen-Shannon distance between $\bm h$ and the probability function $\bm u=(u(1), u(2), \dots, u(D!))$ stems as an adequate measure of such a distance:
\begin{equation}
Q'(\bm{h}, \bm{u}) = \sum_{\ell=1}^{D!} \Big(h(\ell) \log\frac{h(\ell)}{u(\ell)} +
u(\ell) \log\frac{u(\ell)}{h(\ell)}
\Big).
\label{eq:JensenShannon}
\end{equation}
This has been the preferred distance function since its proposal by \citet{Lamberti2004}.
Most works consider the uniform distribution $\bm u = (1/D!, 1/D!, \dots, 1/D!)$ as the reference model.

@@

We illustrate the use of the Entropy-Complexity plane ($H\times C$) with the following time series:
\begin{itemize}
\item Colored $k$-noise, or $f^{-k}$ noise: white ($k=0$), $k=1/2$, pink ($k=1$), $k=3/2$, red ($k=2$), $k=5/2$, and $k=3$;
\item Chaotic logistic series $x_t = r x_{t-1} (1 - x_{t-1})$, with $r=3.6$ and $4$;
\item Deterministic series: monotonic increasing ($\log(x_t+0.1)$, $x_t=\{1,2,\dots,10^4$) and periodic ($\sin(2x_t)\cos(2x_t)$, with $0\leq x_t\leq 2\pi$ over ten thousand equally spaced points).
\end{itemize}
In all cases, we used $D=6$ and $\tau=1$.
Fig.~\ref{fig:Histograms} shows nine of the histograms produced by these series using the Mersenne-Twister pseudorandom number generator;
we omitted those corresponding to the deterministic series, as they produce one and two nonzero bins.

>>>>

\fundinginfo{CNPq, Fapeal, and FAPESP (São Paulo Research Foundation) grant 2020/05121-4.}

@@

Let ${\mathbbm A}_{D}$ (with $D \geq 2$ and $D \in {\mathbbm N}$) be the symmetric group of order $D!$ formed by all 
possible permutation of order $D$, and the symbol component vector 
${\bm \pi}^{(D)} = (\pi_1, \pi_2, \dots, \pi_D)$ so every element ${\bm \pi}^{(D)}$ is unique 
($\pi_j \neq \pi_k$ for every $j \neq k$). 
Consider for the time series ${\mathcal X} \equiv \{x_t\}_{t=1}^{T}$ its time delay embedding representation,
with embedding dimension $D \geq 2$ ($D \in {\mathbbm N}$) and time delay $\tau \geq 1$ ($\tau \in {\mathbbm N}$, also called ``embedding time''):
\begin{equation} 
{\mathbf X}^{(D,\tau)}_t =\big( x_t,x_{t+\tau},\dots,x_{t+(D-1)\tau} \big) ,
\label{eq:time-delay}
\end{equation} 
for $t = 1,2,\dots,N$ with $N = T-(D-1) \tau$.
Then, the vector ${\mathbf X}^{(D,\tau)}_t$ can be mapped to a symbol ${\bm \pi}^{(D)} \in {\mathbbm A}_{D}$. 
This mapping should be defined in a way that preserves the desired relation between the elements 
$x_t  \in {\mathbf X}^{(D,\tau)}_t$, and all $t \in T$ that share this pattern (also called ``motif'') are mapped to the same 
${\bm \pi}^{(D)}$. 
The two most frequent ways to define the mapping ${\mathbf X}^{(D,\tau)} \mapsto {\bm \pi}^{(D)}$ are:  
\begin{enumerate}[label=\alph*)]
\item ordering the ranks of $x_t \in {\mathbf X}^{(D,\tau)}$ in chronological order 
(\textit{Rank Permutation}) or,
\item ordering the time indexes of $x_t \in {\mathbf X}^{(D,\tau)}$  
(\textit{Chronological Index Permutation}).
\end{enumerate}
See details in the work by \citet{BPRepeatedValuesChaos}.
Without loss of generality, in the following, we will use the latter.

@@

After computing all the symbols, one obtains the histogram of proportions $\bm h = \big(h(j)\big)_{1\leq j\leq D!}$.
Such histogram estimates the (unknown, in general) probability distribution function of these patterns.
The next step into the characterization of the time series is computing descriptors from this histogram.

@@

The Jensen-Shannon distance between $\bm h$ and the probability function $\bm u=\big(u(1), u(2), \dots, u(D!)\big)$ stems as an adequate measure of such a distance:
\begin{equation}
Q'(\bm{h}, \bm{u}) = \sum_{\ell=1}^{D!} \Big(h(\ell) \log\frac{h(\ell)}{u(\ell)} +
u(\ell) \log\frac{u(\ell)}{h(\ell)}
\Big).
\label{eq:JensenShannon}
\end{equation}
This has been the preferred distance function since its proposal by \citet{Lamberti2004}.
Most works consider the uniform distribution $\bm u = (1/D!, 1/D!, \dots, 1/D!)$ as the reference model.
In this case, \eqref{eq:JensenShannon} becomes
\begin{equation}
Q'(\bm{h}, \bm{u}) = \sum_{\ell=1}^{D!} h(\ell) \log h(\ell) - \frac{1}{D!} \sum_{\ell=1}^{D!} \log h(\ell).
\label{eq:JS-Uniform}
\end{equation}
The usual assumption in~\eqref{eq:JensenShannon} and~\eqref{eq:JS-Uniform} is $0\log 0=0$.

@@

We illustrate the use of the Entropy-Complexity plane ($H\times C$) with the following time series:
\begin{itemize}
\item Colored $k$-noise, or $f^{-k}$ noise: white ($k=0$), $k=1/2$, pink ($k=1$), $k=3/2$, red ($k=2$), $k=5/2$, and $k=3$;
\item Chaotic logistic series $x_t = r x_{t-1} (1 - x_{t-1})$, with $r=3.6$ and $4$;
\item Deterministic series: monotonic increasing ($\log(x_t+0.1)$, $x_t=\{1,2,\dots,10^4\}$) and periodic ($\sin(2x_t)\cos(2x_t)$, with $0\leq x_t\leq 2\pi$ over ten thousand equally spaced points).
\end{itemize}
In all cases, we used $D=6$ and $\tau=1$.
Fig.~\ref{fig:Histograms} shows nine of the histograms produced by these series using the Mersenne-Twister pseudorandom number generator;
we omitted those corresponding to the deterministic series, as they produce one and two nonzero bins.
